---
title: "day_3_nov_1_2016_occupancy_ttest"
output: pdf_document
---

Today I want to run some t-tests for 2 populations (occupied vs not occupied).

```{r, echo=FALSE}
df <- read.csv("~/Downloads/databank/occupancy_data/datatest.txt")
df$date <- as.POSIXct(df$date)
```

First I will create separate variables for when the office is occupied vs not.

```{r}
occupied <- df[df$Occupancy == 1,]
not_occupied <- df[df$Occupancy == 0,]
summary(occupied)
summary(not_occupied)
```

### Humidity 

We hypothesize that the humidity is different for occupied vs not occupied. Since I have no idea how humidity works, I will just run a 2-tailed test. 

```{r}
hist(occupied$Humidity)
hist(not_occupied$Humidity)
t.test(occupied$Humidity, not_occupied$Humidity)
```

Just eye-balling the p-value, we can tell that it's a very small value. This means that we can confidently reject the null that the humidity is equal amongst occupied vs not occupied.

```{r}
library(lsr)
cohensD(occupied$Humidity, not_occupied$Humidity)
```

Wow I didn't know that cohen's D can go above 1. I usually just see values greater than 0.8. 

It's important to remember that cohen's D doesn't actually have any statistical interpretation, just that the difference in the mean is a lot greater than the pooled standard deviation between the two samples.

