---
title: 'W271 Live Session 10: VAR'
author: "Jeffrey Yau and Devesh Tiwari"
date: "3/10/2017"
output:
  pdf_document: default
  html_notebook: default
---

# Main topics covered in Week 10
    - Regression with multiple trending time series
    - Correlation of time series with trends
    - Spurious correlation
    - Unit-root non stationarity and Dickey-Fuller Test
    - Cointegration
    - Multivariate Time Series Models: Vector Autoregressive (VAR) model
        - Estimation, model diagnostics, model identification, model selection, assumption testing, and statistical inference / forecasting
    - Notion of cross-correlation


# Readings
**CM2009:** Paul S.P. Cowpertwait and Andrew V. Metcalfe. *Introductory Time Series with R*. Springer. 2009. 
  
  - Ch. 11

**SS2016:** Robert H. Shumway and David S. Stoffer. *Time Series Analysis and Applications.* EZ Edition with R Examples:
  
  - None

**HA:** Rob J Hyndman and George Athanasopoulos. Forecasting: Principles and Practice: 
  
  - Ch. 9.2


# Agenda for this week's live session:

  1. Review the steps to build a VAR model: Example 1 - a given series (series2.csv)
  
  2. Review the steps to buld a VAR model: Example 2 - cmort and tempr series

  3. Other topics (if time permit)

Some start-up codes:
```{r}
#sessionInfo()

# Insert the function to *tidy up* the code when they are printed out
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)

# Set Numeric Value Display
# See reference from https://stat.ethz.ch/R-manual/R-devel/library/base/html/options.html
options(digits=2) # Set the printed number of digits to 2. Note: It is a suggestion only. Default is 7. 
#options("scipen" = 10)

# Set memory limit
memory.limit(50000000)

# Clean up the workspace before we begin
rm(list = ls())

# Set working directory
wd <- "~/Documents/JStuff/Teach/w271/LiveSessions/week10" 
setwd(wd)

# Load libraries
library(car)
library(dplyr)
library(Hmisc)

library(forecast)
library(astsa)
library(vars)
```


# 1. Build a VAR model using the given series

  **Task: 1. load the series**
  **      2. Conduct EDA using the series**
  - *Estmated Time: Breakout session 10 minutes*
  - *Estmated Time: Class discussion 10 minutes*

Load the series and examine them.  Consider them as monthly series, spanning over a period of 11 years, starting in 2000. You will practice building a VAR model using these two series.

```{r}
df <- read.csv("~/Downloads/Quick-Data-Science-Experiments-2017/time_series/time_series_course/unit11/series2.csv")

str(df)
s1 <- ts(df$y1, frequency = 12)#, start=c(2000, 1))
s2 <- ts(df$y2, frequency = 12)#, start=c(2000, 1))


plot(s1)
plot(s2)


acf(s1, lag.max = 100)
acf(s2, lag.max = 100)

ccf(s1, s2)
```


Conduct EDA on the series
```{r}
# YOUR CODE TO BE HERE
```

  **Task: 1. Estimate a VAR model using the entire sample series and choose the optimal order using Schwartz Criterion**
  **      2. Conduct model diagnostic **
  **      3. Produce a 3-month forecast **
  - *Estmated Time: Breakout session 15 minutes*
  - *Estmated Time: Class discussion 15 minutes*

```{r}
#############
?VAR

#adf.test(cbind(s1, s2))

m <- VAR(cbind(s1, s2), p=2, type="both")
summary(m)

?VARselect
m_vs <- VARselect(cbind(s1, s2), lag.max = 20,type = "both")
m_vs$criteria
m_vs$selection

serial.test(m, lags.pt = 12, type = "PT.asymptotic" )

#############

plot(m)

summary(m)

res1 <- ts(m$varresult$s1$residuals)
res2 <- ts(m$varresult$s2$residuals)

acf(res1)
acf(res2)

ccf(res1, res2)

?po.test # null is that x is not cointegrated
po.test(cbind(res1, res2))

#############

plot(predict(m, n.ahead = 3))

#############

# 

```


# 2. Build a VAR Model using the cmort and tempr series 

  ** Task: Repeat the steps above except that now you will train the model using the training set and select your model based on RMSE**
  - *Estmated Time: Breakout session 20 minutes*
  - *Estmated Time: Class discussion 15 minutes*

Use the following code to generate your training and test series
```{r}
library(vars)
cmort.train <- cmort[1:458]
cmort.train <- ts(cmort.train, frequency = 52,
                  start = c(1979, 1))
cmort.test <- cmort[459:508]

tempr.train <- tempr[1:458]
tempr.train <- ts(tempr.train, frequency = 52,
                  start = c(1979, 1))
tempr.test <- tempr[459:508]

mortality <- cbind(cmort.train, tempr.train)
```


