---
title: 'W271 Live Session 10: VAR'
author: "Jeffrey Yau and Devesh Tiwari"
date: "3/10/2017"
output:
  pdf_document: default
  html_notebook: default
---

# Main topics covered in Week 10
    - Regression with multiple trending time series
    - Correlation of time series with trends
    - Spurious correlation
    - Unit-root non stationarity and Dickey-Fuller Test
    - Cointegration
    - Multivariate Time Series Models: Vector Autoregressive (VAR) model
        - Estimation, model diagnostics, model identification, model selection, assumption testing, and statistical inference / forecasting
    - Notion of cross-correlation


# Readings
**CM2009:** Paul S.P. Cowpertwait and Andrew V. Metcalfe. *Introductory Time Series with R*. Springer. 2009. 
  
  - Ch. 11

**SS2016:** Robert H. Shumway and David S. Stoffer. *Time Series Analysis and Applications.* EZ Edition with R Examples:
  
  - None

**HA:** Rob J Hyndman and George Athanasopoulos. Forecasting: Principles and Practice: 
  
  - Ch. 9.2


# Agenda for this week's live session:

  1. Review the steps to build a VAR model: Example 1 - a given series (series2.csv)
  
  2. Review the steps to buld a VAR model: Example 2 - cmort and tempr series

  3. Other topics (if time permit)

Some start-up codes:
```{r}
#sessionInfo()

# Insert the function to *tidy up* the code when they are printed out
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)

# Set Numeric Value Display
# See reference from https://stat.ethz.ch/R-manual/R-devel/library/base/html/options.html
options(digits=2) # Set the printed number of digits to 2. Note: It is a suggestion only. Default is 7. 
#options("scipen" = 10)

# Set memory limit
memory.limit(50000000)

# Clean up the workspace before we begin
rm(list = ls())

# Set working directory
wd <- "~/Documents/JStuff/Teach/w271/LiveSessions/week10" 
setwd(wd)

# Load libraries
library(car)
library(dplyr)
library(Hmisc)

library(forecast)
library(astsa)
library(vars)
```


# 1. Build a VAR model using a given series

  **Task: 1. load the series**
  **      2. Conduct EDA using the series**
  - *Estmated Time: Breakout session 10 minutes*
  - *Estmated Time: Class discussion 10 minutes*

Load the series and examine them.  Consider them as monthly series, spanning over a period of 11 years, starting in 2000. You will practice building a VAR model using these two series.

```{r}
df <- read.csv("series2.csv", header = TRUE, sep=",")
  str(df)
  #View(df[,-1])
# Get rid of the first column
df2 <- df[,-1]
  str(df2)

# Make it into a multivariate time series
y.ts <- ts(df2, frequency = 12, start = c(2000,1))
  str(y.ts)
  head(y.ts)
  tail(y.ts)
```


Conduct EDA on the series
```{r}
plot.ts(y.ts, 
        main="Series y1, y2")

# Scatterplot Matrix, which displays the contemporaneous correlation
scatterplot.matrix(~y.ts[,1]+y.ts[,2]);
  title("Contemporaneous Correlation of Series y1, y2, and y3")
  
# ACF and PACF of each of the individual series
par(mfrow=c(2,2)) 
  hist(y.ts[,1], main=""); title("y1")
  plot.ts(y.ts[,1], main=""); title("y1")
  acf(y.ts[,1], main=""); title("ACF of y1")  
  pacf(y.ts[,1], main=""); title("PACF of y1")  

par(mfrow=c(2,2)) 
  hist(y.ts[,2], main=""); title("y2")
  plot.ts(y.ts[,2], main=""); title("y2")
  acf(y.ts[,2], main=""); title("ACF of y2")  
  pacf(y.ts[,2], main=""); title("PACF of y2")  

# Correlation and Cross-correlation between the two series
cat("Correlation Matrix: ", cor(y.ts[,1],y.ts[,2]))  
par(mfrow=c(1,1))
ccf(y.ts[,1],y.ts[,2],main="CCF of the y1 and y2 Series") 
acf
```

Note that it is no longer that case that we can use the individual ACF and PACF to deterimine the order of a VAR(p) models. We, however, still conduct the EDA because we need to know how to set upthe VAR model in R.  

In what follow, we will estimate a series VAR model, using information criterion for the selection. In practice, one could also use out-of-sample forecast performance measurement. We will also conduct model diagnostic and forecasting. We will use the *VARselect()* function to select the optimal model (based on information criterion) (again, for now).


  **Task: 1. Estimate a VAR model using the entire sample series and choose the optimal order using Schwartz Criterion**
  **      2. Conduct model diagnostic **
  **      3. Produce a 3-month forecast **
  - *Estmated Time: Breakout session 15 minutes*
  - *Estmated Time: Class discussion 15 minutes*



```{r}
# Please read the documentation "vars"
# For now, we check out the arguments of the VAR() and VARselect() function

args(VAR)
args(VARselect)

# Select a var model according to information criteria
#roots(, modulus = TRUE)

VARselect(y.ts, lag.max = 5, type = "trend")
    p<-VARselect(y.ts, lag.max = 5, type = "trend")$selection
    p
    
# Estimate the model (based on the order chosen above)
var.mod1 <- VAR(y.ts, p=p[2], type="trend")
  summary(var.mod1)

#################################################################
# Another way to build model
# Split the series into training and test sets

y.train <- y.ts[1:120,]
y.test  <- y.ts[121:length(y.ts[,1]),]
str(y.train)
str(y.test)

y.train<-ts(y.train); str(y.train)
y.test<-ts(y.test); str(y.test)

# Define the Root Mean Square Error (RMSE) function
calculate_rmse <- function(fcast, test){ 
  rmse <- sqrt(mean((fcast - test)^2))
}

# Select optimal order according to out-of-sample forecast error: An example
for (order in 1:5){
  var.mod2 <- VAR(y.train, p = order, type = c("const"))
  var.forecast <- predict(var.mod2, n.ahead = 15)
  var.forecast
  var.forecast$fcst
  y.test
  test.error <- calculate_rmse(var.forecast$fcst$y1[,1], y.test[,1])
  print(c(order, test.error))
} 
```
Typically, after estimating a model, it is a good idea to interpret the estimated coefficients. However, for models with system of equations, such as VAR(p), there is no easy way to interpret the estimated coefficients.

Let's proceed to model diagnostic

The *roots()* function compute the eigenvalues of the companion matrix, and if the system is stable, the eigenvalues should all be less than 1.

Note that the *var()* function consists of 3 types of arguments:
    - the data matrix object y (or an object that can be coerced to a matrix)
    - the integer lag-order p 
    - the type of deterministic regressors to include into the VAR(p).


```{r}  
# Diagnostic testing
  # arch(), normality(), serial(), stability()

  acf(resid(var.mod1)[,1])  
  acf(resid(var.mod1)[,2])

# Normality
 var.mod1.norm <- normality.test(var.mod1, multivariate.only = TRUE)
 names(var.mod1.norm) 
 var.mod1.norm
 
# Serial Correlation
  # Portmanteau Test
  var.mod1.pt.asy <- serial.test(var.mod1, lags.pt = 16, type = "PT.asymptotic")
  var.mod1.pt.asy 

  # Breush-Godfrey LM-Statistics
  serial.test(var.mod1, lags.pt = 16, type = "BG")
  
# Test ARCH effect: To test for heteroscedasticity
# Use univariate and multivariate ARCH test
 args(arch)
 var.mod1.arch <- arch.test(var.mod1)
 names(var.mod1.arch)
 var.mod1.arch

# Forecast for the next 3 months
  var.mod1.f10 <- predict(var.mod1, n.ahead = 3, ci = 0.95)
  class(var.mod1.f10)
  var.mod1.f10  
  plot(var.mod1.f10)
```

# Build a VAR Model using the cmort and tempr series 

  ** Task: Repeat the steps above except that now you will train the model using the training set and select your model based on RMSE**
  - *Estmated Time: Breakout session 20 minutes*
  - *Estmated Time: Class discussion 15 minutes*

Use the following code to generate your training and test series
```{r}
library(vars)
cmort.train <- cmort[1:458]
cmort.train <- ts(cmort.train, frequency = 52,
                  start = c(1979, 1))
cmort.test <- cmort[459:508]

tempr.train <- tempr[1:458]
tempr.train <- ts(tempr.train, frequency = 52,
                  start = c(1979, 1))
tempr.test <- tempr[459:508]

mortality <- cbind(cmort.train, tempr.train)
```


