---
title: 'W271 Live Session 12: Analysis of Panel Data 1'
author: "Jeffrey Yau and Devesh Tiwari"
date: "4/4/2017"
output:
  pdf_document: default
  html_notebook: default
---

# Main topics covered in Week 12 (Async Unit 11)
    - Fixed Effect Model
    - A Digression: differencing when there are more than 2 time periods
    - Random effect model
    - Fixed effect vs. random effect models

# Readings
**W2016:** Jeffrey Wooldridge. *Introductory Econometrics: A Modern Approach.* 6th edition. Cengage Learning 

    - Ch. 14.1 - 14.2
    - [package plm](https://cran.r-project.org/web/packages/plm/plm.pdf)
    - [plm vignettes](https://cran.r-project.org/web/packages/plm/vignettes/plm.pdf)

# Agenda for this week's live session:

  1. Exploratory data analysis of panel (or longitudinal) data
  
  2. Fixed-effect models, Random-effect models, various diagnostic tests

Some start-up codes:
```{r}
#sessionInfo()

# Insert the function to *tidy up* the code when they are printed out
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)

# Set Numeric Value Display
# See reference from https://stat.ethz.ch/R-manual/R-devel/library/base/html/options.html
options(digits=2) # Set the printed number of digits to 2. Note: It is a suggestion only. Default is 7. 
#options("scipen" = 10)

# Set memory limit
memory.limit(50000000)

# Clean up the workspace before we begin
rm(list = ls())

# Set working directory
wd <- "~/Documents/JStuff/Teach/w271/LiveSessions/week13"
setwd(wd)

# Load libraries
library(car)
library(dplyr)
library(Hmisc)
library(ggplot2)
library(lattice)
library(plm)
```


# 1. Exploratory data analysis of panel data

  - *Estmated Time: Breakout session 15 minutes*
  - *Estmated Time: Class discussion 10 minutes*

In this example, we use a modified panel dataset on 7 countries. The *yIndex* is a measure of the well-being of the population in a country, considering that as some health measure. A bigger number reflects a better health status. The *xIndex* variable is a (normalized) measure of the economic activities of the country; a bigger number indiciates "better" economic performance. *country* indicates, well, country, and *year* indicates the year in which the health and economic measurements were taken. The question of interest is "Does better economic activities, as measured by *xIndex*, improve the country's health status, as measured by *yIndex*?"

**Tasks:** 
    - Import the data as data.frame. 
    - Examine the basic structure of each of the datasets. 
    - Try to understand each of the variables
    - Discuss about the dataset
    - At what level is this dataset?
    - Conduct a throughout EDA, including a growth curve analysis that was covered in the live session in week 12
    
```{r}
# YOUR CODE HERE
df <- read.csv("~/Downloads/Quick-Data-Science-Experiments-2017/time_series/time_series_course/unit13/data_wk13.csv")

head(df)
summary(df)

sort(unique(df$year))
sort(unique(df$country))
sort(unique(df$yIndex))
sort(unique(df$xIndex))

hist(df$yIndex)

#percentile(df$xIndex, 20)

scatterplot(df$xIndex, df$yIndex)
xyplot(yIndex ~ year | country, data=df, as.table=T)
xyplot(xIndex ~ year | country, data=df, as.table=T)
xyplot(yIndex ~ xIndex | country, data=df, as.table=T)

ggplot(data=df[df$country=='C',], aes(x=year, y=yIndex)) +
    geom_bar(stat="identity")


library(foreign)
library(gplots)

coplot(yIndex ~ year|country, type="l", data=df)

coplot(xIndex ~ year|country, type="b", data=df)

plotmeans(xIndex ~ country, main="Heterogeineity of an Economic Index across countries", data=df)
plotmeans(yIndex ~ country, main="Heterogeineity of a Health Index across countries", data=df)

```


# 2. OLS Regression Model

  - *Estmated Time: Breakout session 10 minutes*
  - *Estmated Time: Class discussion 5 minutes*

**Tasks:** 
  - Suppose you want to answer the question posted above using an OLS regression model.
  - Is OLS regression model a reasonable approach? Why? Why not?
  - Don't forget to interpret the result! So, according to your model, what's your answer to the question posted above?
  - What do you have to assume to make this approach reasonable?
  - Are these assumptions that you will have to make valid?
  

```{r}
# yIndex is ordinal
# omitted variable bias
# SE is not significant --> not independent observations (serial correlation), data is across time
m <- lm(yIndex ~ xIndex, data=df)
plot(m)
```

# 3. Fixed-Effect Regression Model

  - *Estmated Time: Breakout session 15 minutes*
  - *Estmated Time: Class discussion 10 minutes*

**Tasks:**
  - Suppose you want to answer the question posted above using a Fixed-Effect regression model.
  - First, write done a fixed-effect regression model with an unobserved effect.
  - What assumptions are needed for the fixed-effect regression model?
  - Explain what a fixed effect is?
  - Think about why a fixed-effect framework would make sense for this question?
  - Estimate a fixed-effect model.
  - In your fixed-effect model, how many groups, time periods, and observations are there?
  - Don't forget to interpret the result! So, according to your model, what's your answer to the question posted above?

```{r}
# y_it = X_it B + \alpha_i + u_it 
# assumption: 1) X_it can be correlated within an obj 2) there is variation within X_it
# fixed effect: fixed \alpha_i ?
# why a fixed effect: can't measure \alpha_i, solves omitted var bias
# y_it_demeaned = x_it_demeaned B + u_it_demeaned
# 
(fe_model <- plm(yIndex ~ xIndex, data=df, index=c("country", "year"), na.action=na.omit, model="within"))
summary(fe_model)
```

# 4. Random-Effect Regression Model

  - *Estmated Time: Breakout session 15 minutes*
  - *Estmated Time: Class discussion 10 minutes*

**Tasks:**
  - Suppose you want to answer the question posted above using a Random-Effect regression model.
  - First, write done a random-effect regression model with an unobserved effect.
  - What assumptions are needed for the random-effect regression model?
  - Explain what a random effect is?
  - Think about why a random-effect framework would make sense for this question?
  - Estimate a fixed-effect model.
  - Don't forget to interpret the result! So, according to your model, what's your answer to the question posted above?


```{r}
# The random effects assumption (made in a random effects model) is that the individual specific effects are uncorrelated with the independent variables
# y_it = \mu + U_i + w_it

(re_model <- plm(yIndex ~ xIndex, data=df, index=c("country", "year"), na.action=na.omit, model="random"))
summary(re_model)
```

# 5. Diagnostic Tests and Final Discussion

  - *Estmated Time: Breakout session 10 minutes*
  - *Estmated Time: Class discussion 5 minutes*

**Tasks:**
  - Conduct diagnostic tests to determine whether or not a fixed effects model or a random effects model is appropriate.
  - Discuss situations (and assumptions) under which fixed effect is more appropriate and situations under

  - Hausman test: null hypothesis is that the preferred model is random effects vs. the alternative the fixed effects. It tests whether the unique errors $u_i$ are correlated with the explanatory variables; the null hypothesis is they are not.
    - Estimate a fixed effects model and a random effect model. Then perform the test. If the p-value is significant (e.g. $p \leq 0.05$), then use fixed effects; otherwise, use random effects.

```{r}
# YOUR CODE HERE
?phtest
phtest(fe_model, re_model)
```


